---
title: "Reproducible Research - Week 2: Course Project 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load and preprocess the data

1. Load the data
```{r, message=FALSE, warning=FALSE}
## loads packages
library(dplyr)
library(lattice)
```


```{r, results='hide'}
##  data file is stored in the working directory
steps <- read.csv("activity.csv")

```


```{r }
## view properties of dataset
str(steps)
```

2. Transform the data 
```{r }
## converts date variable to date class
steps$date <- as.Date(steps$date, "%Y-%m-%d")
```


## What is mean total number of steps taken per day?

For this part of the assignment, the missing values in the dataset are ignored.

1. Calculate the total number of steps taken per day
```{r, message=FALSE}
## uses dplr package to calculate and summarize total daily steps
stepsdailytotal <-
  steps %>% 
  group_by(date) %>%        
  summarize(total= sum(steps, na.rm = TRUE))

head(stepsdailytotal)

```


2. Make a histogram of the total number of steps taken each day
```{r}
## histogram of total steps per day 
hist(stepsdailytotal$total,
     xlab = "Steps",
     main = "Total Steps per Day")
```

3. The mean and median of the total number of steps taken per day

```{r}
paste("Mean is ", round(mean(stepsdailytotal$total, na.rm = TRUE), 0)) 
paste("Median is ", round(median(stepsdailytotal$total, na.rm = TRUE), 0)) 
```


## What is the average daily activity pattern?

1. A time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

```{r, message=FALSE}
## calculates averages
pattern <-
  steps %>% 
  group_by(interval) %>%        
  summarize(avg= mean(steps, na.rm = TRUE))
```


```{r, message=FALSE,warning=FALSE}
head(pattern)
```


```{r}
## plots averages
plot(pattern$interval,
     pattern$avg, 
     type="l", 
     ylab = "Average steps",
     xlab = "Interval",
     main = "Daily Activity Pattern" )
```

2. The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps.

```{r}

paste("Interval ", 
      pattern$interval[pattern$avg==max(pattern$avg)], 
      " contains the maximum number of steps") 

```

## Imputing missing values

1. The total number of missing values in the dataset (i.e. the total number of rows with \color{red}{\verb|NA|}NAs) is 2304.

```{r}
colSums(is.na(steps))
```

2. Strategy for filling in all of the missing values in the dataset is to impute the mean for that 5-minute interval.

Subset missing values

```{r}
## subsets rows with missing steps value
missing <- is.na(steps$steps)
missingsubset <- steps[missing,]
```


```{r}
## confirms only NAs are subsetted
head(missingsubset)
tail(missingsubset)
```


Replace missing values with interval average

```{r}
## replaces missing values with  average calculated in Section 3 
for (i in 1:length(missingsubset$steps))  {
  missingsubset[i,1] <- pattern$avg[pattern$interval==missingsubset$interval[i]]
  }
```


```{r}
str(missingsubset)
```


```{r}
## confirms no missing data in subset
colSums(is.na(missingsubset))
```

3. Creates a new dataset that is equal to the original dataset but with the missing data filled in

```{r}
## creates copy of original dataset and inserts missing values
stepsimputed <- steps
stepsimputed[missing,] <- missingsubset
```


```{r}
## confirms no missing values
colSums(is.na(stepsimputed))
```

4. Histogram of the total number of steps taken each day

```{r, warning=FALSE, message=FALSE}
## recalculates daily total with imputed missing values 

stepsfilled <-
  stepsimputed %>% 
  group_by(date) %>%        
  summarize(total= sum(steps, na.rm = TRUE))
```


```{r, warning=FALSE}
## plots data with imputed values included
hist(stepsfilled$total, 
     xlab = "Steps",
     main = "Total Steps per Day")
```

The mean and median total number of steps taken per day


```{r}
paste("Mean is ", round(mean(stepsfilled$total), 0)) 

paste("Median is ", round(median(stepsfilled$total),0)) 
```


Median and mean are now identical and both values are higher than the previous estimates. 


## Are there differences in activity patterns between weekdays and weekends?

Dataset with the filled-in missing values is used for this part.

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}

stepsimputed$day <- weekdays(steps$date)
stepsimputed$dayfac <- "Weekday"
weekendInd <- which(stepsimputed$day=="Saturday" |  stepsimputed$day =="Sunday")
stepsimputed$dayfac[weekendInd] <- "Weekend"
```

Panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 


```{r, warning=FALSE, message=FALSE}
stepsdays <-
  stepsimputed %>% 
  group_by(dayfac, interval) %>% 
  summarize(avg= mean(steps, na.rm = TRUE))
```


```{r, warning=FALSE, message=FALSE}
xyplot(avg ~ interval | dayfac, 
       data = stepsdays, 
       layout = c(1, 2),
       type="l")
```


